---
title: "Conditional Chunk Execution in R Markdown"
date: 2021-01-30T17:14:59+00:00
categories: ["notes"]
tags: ["R", "R Markdown"]
summary: "example of using params to toggle eval chunk option with R Markdown"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
catn <- function(x="") cat("    ", x, "\n")
```

When generating documents with R Markdown, it may be useful to conditionally evaluate certain chunks programmatically. 

One way to do so is to use a combination of two `knitr` and `rmarkdown` features:

- Parameters[^parameter-doc]
- The "eval" chunk option[^eval-doc]

[^parameter-doc]: https://rmarkdown.rstudio.com/lesson-6.html
[^eval-doc]: https://bookdown.org/yihui/rmarkdown/r-code.html

The `rmarkdown::render()` function accepts a list of parameters to pass into the document while rendering. When written with consideration, these parameters can toggle conditions to be passed to the "eval" chunk option.

The example below is a minimal demonstration of the workflow.

## Example

First, we need to write the R Markdown document. In this case, the rendered output will be a PDF conditionally created with the following scenarios:

- Both a plot and a table
- Just a plot
- Just a table
- Neither a plot nor a table

The parameters passed in will be logicals for "plot" and "table". These values are used to establish conditions in a chunk near the top of the R Markdown file.

The full `.Rmd` file:

```{r, results="asis"}
catn('---')
catn('title: "Sleep Summary"')
catn('output: pdf_document')
catn('params:')
catn('  plot: TRUE')
catn('  table: TRUE')
catn('---')
catn('')
catn('```{r setup, include=FALSE}')
catn('knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)')
catn('library(tidyverse)')
catn('')
catn('## establish conditions')
catn('eval_plot <- params$plot')
catn('eval_table <- params$table')
catn('eval_both <- eval_plot & eval_table')
catn('eval_one_of <- eval_plot | eval_table')
catn('eval_neither <- !eval_one_of')
catn('```')
catn('')
catn('```{r, prep, eval=eval_one_of}')
catn('knitr::asis_output("## Summary")')
catn('')
catn('## summarize msleep data')
catn('msleep_summary <-')
catn('  msleep %>%')
catn('  group_by(order) %>%')
catn('  summarise(sleep = round(mean(sleep_total),2), .groups = "drop") %>%')
catn('  arrange(desc(sleep))')
catn('```')
catn('')
catn('```{r, table-heading, eval=eval_table}')
catn('knitr::asis_output("### Table")')
catn('```')
catn(' ')
catn('```{r, table, eval=eval_table}')
catn('knitr::kable(msleep_summary)')
catn('```')
catn('')
catn('```{r, plot-heading, eval=eval_plot}')
catn('knitr::asis_output("### Plot")')
catn('```')
catn('')
catn('```{r, plot, eval=eval_plot}')
catn('msleep_summary %>%')
catn('  ggplot(aes(reorder(order,sleep), sleep)) +')
catn('  geom_col() +')
catn('  coord_flip() +')
catn('  labs(x = NULL, y = "Sleep total (Hours)") +')
catn('  theme_minimal()')
catn('```')
catn('')
catn('```{r, empty-msg, eval=eval_neither}')
catn('knitr::asis_output("## WARNING: Neither plot nor table generated ...")')
catn('```')
catn('')
```

Assuming the content above is stored in a file called `example.Rmd` we can trigger the processing for each condition.

To generate a PDF with both the plot and the table:

```{r, echo=TRUE, eval=FALSE}
rmarkdown::render("example.Rmd", 
                  output_file = "example-both.pdf", 
                  params = list(plot = TRUE, table = TRUE))
```

<object data="/img/example-both.pdf" type="application/pdf" width="700px" height="700px">
  <embed src="/img/example-both.pdf">
  <p>This browser does not support PDFs. Download the PDF to view it: <a href="/img/example-both.pdf">Download PDF</a>.</p>
  </embed>
  </object>
  
To generate a PDF with just the plot:

```{r, echo=TRUE, eval=FALSE}
rmarkdown::render("example.Rmd", 
                  output_file = "example-plot.pdf", 
                  params = list(plot = TRUE, table = FALSE))
```

<object data="/img/example-plot.pdf" type="application/pdf" width="700px" height="700px">
  <embed src="/img/example-plot.pdf">
  <p>This browser does not support PDFs. Download the PDF to view it: <a href="/img/example-plot.pdf">Download PDF</a>.</p>
  </embed>
  </object>
  
To generate a PDF with just the table:

```{r, echo=TRUE, eval=FALSE}
rmarkdown::render("example.Rmd", 
                  output_file = "example-table.pdf", 
                  params = list(plot = FALSE, table = TRUE))
```

<object data="/img/example-table.pdf" type="application/pdf" width="700px" height="700px">
  <embed src="/img/example-table.pdf">
  <p>This browser does not support PDFs. Download the PDF to view it: <a href="/img/example-table.pdf">Download PDF</a>.</p>
  </embed>
  </object>
  
To generate a PDF with neither:

```{r, echo=TRUE, eval=FALSE}
rmarkdown::render("example.Rmd", 
                  output_file = "example-neither.pdf", 
                  params = list(plot = FALSE, table = TRUE))
```

<object data="/img/example-neither.pdf" type="application/pdf" width="700px" height="700px">
  <embed src="/img/example-neither.pdf">
  <p>This browser does not support PDFs. Download the PDF to view it: <a href="/img/example-neither.pdf">Download PDF</a>.</p>
  </embed>
  </object>
