---
title: "Looping Across Files"
date: 2019-06-29T17:14:59+00:00
categories: ["notes"]
tags: ["R", "bash"]
summary: "comparing bash and R for iterative file manipulation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

When performing repetitive file manipulation operations, it can be useful to write a loop. 

Let's assume you have a file structure organized as follows:

```
.
+-- loop.sh
+-- loop.R
+-- raw
|   +-- sample1.bed
|   +-- sample2.bed
|   +-- sample3.bed
|   +-- sample4.bed
|   +-- sample5.bed
+-- processed
|   +-- 
```

You'd like to loop through all of the `.bed` files in the `raw/` directory, and output the first and last columns in a `.csv` file in the `processed/` directory that shares the same file prefix (i.e. "sample1") as the given input. 

To do so you could write a bash script that is in a file called `loop.sh`:

```{bash}
#!/bin/bash

# loop through all .bed files in the raw dir
FILES=raw/*.bed

for f in $FILES

do
  # message
  printf "\n%s\n%s\n" "processing $f ..." "------------------------------"

  # get input file name ...
  # ... and create output file based on that
  in_fn=$(basename -- "$f")
  out_fn="${in_fn%.*}"

  # use awk to get first and last columns  ...
  # ... pipe output to csv format
  awk 'BEGIN{FS="\t";OFS=","}{print $1, $NF}' $f > processed/"$out_fn".csv

done
```

And in `loop.R` you could write something like this:

```{r}
files <- list.files("raw", full.names = TRUE)

for (f in files) {
  
  bed <- read.delim(f, sep = "\t")
  
  out_fp <- paste0("processed/", 
                   tools::file_path_sans_ext(basename(f)),
                   ".csv")
  
  write.csv(bed[,c(1,ncol(bed))], row.names = FALSE, file = out_fp)
  
}
```

Both of the above should work and yield the following: 

```
.
+-- loop.sh
+-- loop.R
+-- raw
|   +-- sample1.bed
|   +-- sample2.bed
|   +-- sample3.bed
|   +-- sample4.bed
|   +-- sample5.bed
+-- processed
|   +-- sample1.csv
|   +-- sample2.csv
|   +-- sample3.csv
|   +-- sample4.csv
|   +-- sample5.csv
```

However, your choice of method could have serious performance implications.

I tested this on dummy files that were roughly 100MB each (~ 2.5 million rows, 4 columns). 

The R method took ~ 90 seconds ... the bash script finished in 30 seconds.

