---
title: "LaTeX Symbols Inside Pandoc Table Column Headings Inside RMarkdown Documents"
date: 2016-03-03T11:26:40-05:00
categories: ["notes"]
tags: ["R Markdown", "R", "LaTex", "Pandoc"]
summary: "RMarkdown(Pandoc(LaTeX))"
---

I just spent ~ 30 minutes figuring out how to include LaTeX symbols ... in a Pandoc table column header ... in an RMarkdown document. 

There are plenty of LaTeX cheat sheets like [this one](https://wch.github.io/latexsheet/latexsheet-0.png) to choose from. And there are plenty of examples of [using LaTeX in RMarkdown](http://www.calvin.edu/~rpruim/courses/m343/F12/RStudio/LatexExamples.html) as well. 

The thing is that the syntax for these LaTeX symbols gets trickier when run inside a chunk (i.e. block of R code). The code below demonstrates one way to get the pander table to output the column names with LaTeX syntax. 

```{r, message = FALSE, warning = FALSE, eval = FALSE}
# set random seed
set.seed(123)

# set K for number of replications
K <- 1000

# rate (lambda)
lambda <- 3

# n 
n <- c(5, 10, 15, 20, 25)

# simulate data
sims <- lapply(n, function(x) replicate(K, rexp(x, lambda)))

# calculate sample means
sims_mean <- lapply(sims, function(x) apply(x, 2, mean))

# calculate sample variances
sims_var <- lapply(sims, function(x) apply(x, 2, var))


library(pander)
xbars <- sapply(sims_mean, mean)
vars <- sapply(sims_mean, var)
expected_s2 <- sapply(sims_var, mean)

propdata <- data.frame(n = n,
                       xbars = xbars, 
                       vars  = vars,
                       expected_s2 = expected_s2)

# assign names of dataframe with latex syntax 
names(propdata) <- c("n", "$\\mu_{\\bar{x}}$", "$\\sigma^2_{\\bar{x}}$", "$E(s^2)$")

# names(propdata) <- c("n", "$mu_", "$\\bar", "$\\sigma")
pander(propdata)
```

The key line is:
  
```{r, eval = FALSE}
names(propdata) <- c("n", "$\\mu_{\\bar{x}}$", "$\\sigma^2_{\\bar{x}}$", "$E(s^2)$")
```

That piece of code assigns names to the `propdata` object the desired syntax ... err ... sort of the desired syntax. You have to escape the `\` character that preceeds each of the references to `mu`, `bar` and `sigma`. 

It's not pretty but it works.

Maybe there's a better way to this with `pandoc.table()`?

