---
title: "pandoc.list"
date: 2016-01-15T14:07:29-05:00
categories: ["notes"]
tags: ["pander", "R markdown"]
summary: "formatting a character vector as a list in R markdown"
---

One of my recent tortures has been formatting RMarkdown reports. The [pander](https://cran.r-project.org/web/packages/pander/index.html) package gives you access to all the [Pandoc options](http://pandoc.org/README.html) you can handle. Seems like a big mountain to climb, but once you get started it's actually not too rough.

In particular, the `pandoc.[WHATEVER]()` functions from pander make it easy to render nicely formatted objects in RMarkdown. 

An example ... say you had the following (not so arbitrary) character vector:


```{r}
bowie <- c("Young Americans", "Station To Station", "Low", "Heroes", "Lodger", "Scary Monsters", "Let's Dance")
```

You could just print the object:

```{r}
bowie
```
Doesn't look great. Maybe try it with the generic `pander()` function around the name of the character vector:

```{r}
# install.packages("pander")
library(pander)
pander(bowie)
```

Still not great, but until a couple days ago that's probably where I would have left it. 

Then I tab-completed into the `pandoc.[WHATEVER]()` functions and realized I was doing it wrong ... 

In this case if I want a list then I should go with `pandoc.list()`:


```{r}
pandoc.list(bowie)
```
That looks more like a list ... at least as it would be printed in a console.

If you want to see it bulleted just wrap another `pander()` around it:

```{r}
pander(pandoc.list(bowie))
```

And if you want to see the vector ordered:

```{r}
pander(pandoc.list(bowie, style="ordered"))
```
The best part about these functions is that they're documented inside R man pages (as opposed to only in the generic pandoc documentation) ... check it out:


```{r eval = FALSE}
?pandoc.list
```

```{r}
sessionInfo()
```
