---
title: "ROC Curves in R"
date: 2019-06-30T17:14:59+00:00
categories: ["notes"]
tags: ["R", "ggplot2"]
summary: "methods for plotting ROC curves"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

> A receiver operating characteristic curve, or ROC curve, is a graphical plot that illustrates the diagnostic ability of a binary classifier system as its discrimination threshold is varied.[^1]

There's already plenty of available information on how to plot ROC curves in R:

- <https://blog.revolutionanalytics.com/2016/08/roc-curves-in-two-lines-of-code.html>
- <https://cran.r-project.org/web/packages/plotROC/vignettes/examples.html>
- <https://campus.datacamp.com/courses/machine-learning-with-tree-based-models-in-r/boosted-trees?ex=12>
- <https://www.youtube.com/watch?v=qcvAqAH60Yw>

A 2019 RViews post compares different methods side-by-side:

<https://rviews.rstudio.com/2019/03/01/some-r-packages-for-roc-curves/>

The example that follows provides a documented method I have used to plot ROC curves, both with the  `pROC` package alone ... and also using data from the `pROC` ROC AUC object and `ggplot2`.

First, some code for to prepare the data (the `Titanic` dataset in this case) for modeling:

```{r}
library(dplyr)

expand_counts <- function(.data, freq_col) {
  
  quo_freq <- dplyr::enquo(freq_col)
  
  freqs <- dplyr::pull(.data, !!quo_freq)
  
  ind <- rep(seq_len(nrow(.data)), freqs)
  
  # Drop count column
  .data <- dplyr::select(.data, - !!quo_freq)
  
  # Get the rows from x
  .data[ind, ]
  
}

titanic <-
  as.data.frame(Titanic, stringsAsFactors = FALSE) %>%
  expand_counts(Freq) %>%
  mutate(Survived = ifelse(Survived == "Yes", 1, 0))

head(titanic)
```

The model will predict survival (yes/no) from the Titanic. Predictors will include class, sex, and age. We'll look at a model of with passenger class as the only predictor versus a model that includes class, sex, and age.

####  `Survived ~ Class`
```{r}
library(pROC)

fit1 <- glm(Survived ~ Class, data = titanic, family = binomial)

prob <- predict(fit1,type=c("response"))

fit1$prob <- prob

roc1 <- roc(Survived ~ prob, data = titanic, plot = FALSE)

roc1
```

####  `Survived ~ Class + Sex + Age`

```{r}
fit2 <- glm(Survived ~ Class + Sex + Age, data = titanic, family = binomial)

prob <- predict(fit2,type=c("response"))

fit2$prob <- prob

roc2 <- roc(Survived ~ prob, data = titanic, plot = FALSE)

roc2
```

```{r}
plot(roc1, lty = "solid")
lines(roc2, lty = "dotted")
```

```{r}
library(ggplot2)

df1 <- 
  data_frame(
    sensitivity = roc1$sensitivities,
    specificity = roc1$specificities,
    thresholds = roc1$thresholds,
    model = "Survived ~ Class"
  )

df2 <- 
  data_frame(
    sensitivity = roc2$sensitivities,
    specificity = roc2$specificities,
    thresholds = roc2$thresholds,
    model = "Survived ~ Class + Sex + Age"
  )

rbind(df1,df2) %>%
  ggplot(aes(1-specificity, sensitivity)) +
  geom_line(aes(group = model, lty = model)) +
  theme(legend.position = "bottom", 
        legend.title = element_blank())
```

[^1]:  From [Receiver operating characteristic](https://en.wikipedia.org/wiki/Receiver_operating_characteristic) by [Wikipedia](https://en.wikipedia.org/) licensed under [CC BY-SA 3.0](https://creativecommons.org/licenses/by-sa/3.0/us/).