---
title: "Include Images in R Markdown PowerPoint"
date: 2020-05-03T12:19:11-04:00
categories: ["notes"]
tags: ["R Markdown", "R"]
summary: "render an entire directory of image files in rmarkdown powerpoint output"
---

I learned about the R Markdown `powerpoint_presentation` output format at [rstudio::conf 2019](http://www.nagraj.net/notes/rstudioconf2019/#rmdpp).

Recently I used the feature to organize a set of previously generated plots (saved as `.png` files) into a single PowerPoint, which was the output format of choice for one collaborator.

Including a single image file in an R Markdown `powerpoint_presentation` is the same as any other Markdown document:

```md
![alt_text](path_to_image.extension)
```

However, including many image files (in my case dozens) would involve tediously repeating that pattern. This post details how to create a loop that generates the code to render every image in a given directory as its own slide in a PowerPoint presentation.

## Example

First we will create example images. This demo uses the `msleep` dataset from **ggplot2**. Here we are generating the images in a separate R script and then writing them to `.png` files to be rendered together in a single output document via R Markdown. 

Note that this is a contrived example. If we were truly starting from scratch, we could use the code below to render the plots directly in R Markdown without saving them as `.png` files at all. However, for my original use-case I already had the image files generated. 


```{r eval = FALSE, echo = TRUE}
library(tidyverse)

## create a list of msleep tibbles split by vore
by_vore <- group_split(msleep, vore)

## make a dir to write the plots
vp_dir <- "vore-plots"
if(!dir.exists(vp_dir)) dir.create(vp_dir)

## loop over each vore and plot the sleep total by bodywt
for(vore in by_vore) {
  
  vore_name <- unique(vore$vore)
  
  ## create the ggplot2 object
  p <-
    ggplot(vore, aes(sleep_total, bodywt)) +
    geom_point() +
    scale_y_log10() +
    ggtitle(label = sprintf("Sleep by body weight (%s-vore)", vore_name)) +
    theme_bw()
  
  print(p)
  
  ## save the plot to a disk
  ggsave(p, filename = sprintf("%s/%s.png", vp_dir, vore_name))
}

```

Now that we have the plots saved to disk, we can create the R Markdown document. This file (here named `vore_plots.Rmd`) will use the `output: powerpoint_presentation` option in the YAML frontmatter. In the body of the `.Rmd` we need a code chunk to generate the raw Markdown text for adding each image to its own slide. This chunk needs to have the `results='asis'` option set for the text issued from `cat()` to render correctly:

```text
    ---
    title: 'vore plots'
    output: powerpoint_presentation
    ---
    
    `r ''````{r results='asis'}
    vps <- list.files('vore-plots', full.names = TRUE)
    
    for(vp in vps) {
      cat('##')
      cat('\n')
      cat(paste0('![](', vp, ')'))
      cat('\n')
      cat('\n')
    }
    ```
```

To generate to the `.pptx` output use the *Knit* button in the RStudio IDE or call `rmarkdown::render()` at the console:

```{r, eval = FALSE, echo = TRUE}
rmarkown::render("vore_ppt.Rmd")
```


