---
title: "iqr4tab and prop4tab"
date: 2020-03-18T17:14:59+00:00
categories: ["notes"]
tags: ["R"]
summary: "helpers for summary statistics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Below are a couple of functions I use for summarizing variables. These are useful for evaluating characteristics in different strata.

## `iqr4tab()`

This function calculates the median and 25th, 75th quartile (IQR) and formats the output for a table:

```{r}
iqr4tab <- function(x, digits = 2, na.rm = TRUE) {
  
  nums <- fivenum(x, na.rm = na.rm)[2:4]
  nums <- round(nums, digits)
  paste0(nums[2], " (", nums[1], ",", nums[3], ")")
  
}
```

## `prop4tab()`

This function will calculate count and proportion for both levels of a binary variable and format for a table:

```{r}
prop4tab <- function(x, level = "1", digits = 1) {
  
  x <- as.factor(x)
  n_total <- length(x)
  
  if (is.na(level) | level == "NA") {
    
    level <- "NA"
    n_level1 <- length(x[x==level])
    
  } else {
    
    n_level1 <- length(x[x==level & !is.na(x)])
    
  }

  prop <- n_level1/n_total
  paste0(n_level1, " (", round(prop*100,digits),")")

}
```

## Example

The following example uses data from the `AmesHousing` package:

```{r}
library(AmesHousing)
library(dplyr)
library(tidyr)

ames <- make_ames()
```

```{r}
ames %>%
  mutate(House_Style = ifelse(House_Style == "One_Story","single story", "multi story")) %>%
  group_by(House_Style) %>%
  summarise(Price = iqr4tab(Sale_Price))
```

```{r}
ames %>%
  mutate(House_Style = ifelse(House_Style == "One_Story","single story", "multi story")) %>%
  group_by(House_Style) %>%
  summarise(Central_Air = prop4tab(Central_Air, level = "Y"))
```

```{r}
ames %>%
  mutate(House_Style = ifelse(House_Style == "One_Story", "single story", "multi story"),
         Fireplace = ifelse(Fireplaces == 0, 0, 1)) %>%
  group_by(House_Style) %>%
  summarise(Price = iqr4tab(Sale_Price),
            Rooms = iqr4tab(TotRms_AbvGrd),
            AC = prop4tab(Central_Air, level = "Y"),
            Fireplace = prop4tab(Fireplace)) %>%
  gather(variable, summary, -House_Style) %>%
  spread(House_Style, summary)
```

